---
title: "Project Proposal"
subtitle: "Tidy Team"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data 

```{r load data}

```

## One Sentence High-Level Goal

Create an interactive spatio-temporal visualization of worldwide deaths related to air pollution. 

## Description of our Goals and Motivation 

A one to two paragraph description of your goals, including your motivation. 
- Depending on the focus of your project, the following might go in here.
- If using a particular dataset(s), a brief description of each dataset including the reason why you chose the particular dataset, its dataset, its dimensions and any other relevant metadata. (Make sure to load the data and use inline code for some of this information.)
- If answering a particular research question, the question itself and the reason why you chose this question.

## Weekly Plan of Attach 

A weekly “plan of attack” outlining your steps to complete your project and including the team member(s) assigned to that task.

## Describe Repo 

The final organization of your project repository. This means describing the project organization in your proposal as well as implementing that organization in your repo. Create any folders needed and place a README.md in each folder explaining what goes in there
