---
title: "Global Deaths by Risk Factor"
subtitle: "A shiny app hosting interactive spatio-temporal visualizations. "
author: "Tidy Team  <br> Drew Flanagan, Courtney Lee, Kathryn Thomas, Raghav Rasal"
institute: "Duke University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
# Remove any packages from this list that you're not using
library(tidyverse)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%"
  )
```

```{r load-data, include=FALSE}
# Load your data here
```

class: inverse, middle, center

# Motivation + Data




- Aim and motivation: Communicate the harmful effects of certain risk factors on 
the world through time via a Shiny app

- 3 main data sets:

  - The World Pollution Data from Kaggle ()
  
  - The World Map Data from Maps Package ()
  
  - Pollution Data from Our World in Data

---

# The Shiny App

### try to hyperlink this directly to the shiny app

---

## Approach

- Data cleaning and merging

 - Utilizing Rds files 
  
- Map with a time functionality 
 - Plotly

- Line plots 

---

class: inverse, middle, center

## Trends and limitations 


---

class: inverse, middle, center

# Thank you for listening!



