---
title: "Air Pollution Deaths"
subtitle: "A shiny app hosting interactive spatio-temporal visualizations. "
author: "Tidy Team  <br> Drew Flanagan, Courtney Lee, Kathryn Thomas, Raghav Rasal"
institute: "Duke University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
# Remove any packages from this list that you're not using
library(tidyverse)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%"
  )
```

```{r load-data, include=FALSE}
# Load your data here
```

class: inverse, middle, center

# Motivation

---

## Inspiration + Data

- Data from Our World in Data
--

- GOAL: The goal of our Shiny App is to communicate the harmful effects of air pollution on the world through time by visually displaying the deaths caused by air pollution and other risk factors.

---

## Workflow

- RMD, for trial and error of plots
--

- Then, Shiny

---

class: inverse, middle, center

# The Shiny App

### try to hyperlink this directly to the shiny app

